Spatiotemporal Multi-Resolution Approximation of the Amari Type Neural Field Model

### 2012-06-11 Multiresolution_framework/ltx/manuscript.tex
minor edits

Neural field models describe the mass action of the central nervous system and are a critical link in our understanding of the biophysics leading to the generation of  electroencephalography (EEG) and magnetoencephalography (MEG) signals. The key components of these models typically describe the macroscopic dynamics of the human brain, but can also be descriptive of finer-scale neurodynamics. Examples are the post-synaptic response kernels (\vik{pulse to wave conversion}) and the activation function (\vik{wave to pulse conversion}) \citep{Jirsa1996,Jirsa1997}. The former can be descriptive of inhibitory and excitatory post-synaptic potentials \vik{in populations of microscopic single neural models, such as integrate and fire neurons,} or an ensemble response to stimulation in macroscopic neural mass models \citep{Stefanescu2008}. The latter can be used to model spiking statistics across spatial scales from single neurons (where it can be considered a cumulative distribution function (CDF) of the probability of firing) \vik{as} mean firing rates of neural masses. The key factor linking the functions of \vik{neural mass models} to particular spatial scales is \vik{the connectivity}, hence the ability to infer a multi-resolution connectivity kernel will enable the formulation of a multi-resolution neural field model.

@@

In this paper, $4^{th}$ -- order cardinal B-spline scaling and wavelet functions are used, resulting in  $8^{th}$ and $12^{th}$ order B-spline functions when calculating Eq.~\eqref{eq:Lambdax} and Eq.~\eqref{eq:Uij} respectively  whose values at integer points are given in Table~\ref{table:MRA-BsplineatIntegerPoints}.

@@

\vik{viktor suggested to delete this part: It is important to point out there is no doubt that there will always be a mismatch between the model and actual cortical tissue. Nevertheless,} the simplifications used in formulating the equations enable the development of new methods for extracting information from electrophysiological data. The trade-off between parsimony and complexity is always difficult when considering neurodynamics and this is further complicated when using estimation algorithms that have huge computational demands with such high-dimensional systems.

@@

By assuming a linear activation function of the form
\begin{equation}
	f(v_t(\mathbf{r})) = \varsigma v_t(\mathbf{r}),
\end{equation}
the state transition equation~\eqref{eq:ApproxDiscreteTimeModel4} simplifies to
\begin{equation}\label{eq:ApproxDiscreteTimeModel_Linear}
	\mathbf{x}_{t+1} = 
	\xi \mathbf{x}_t + 
	\varsigma T_s \mathbf{\Lambda}_{x}^{-1} \int_{\Omega}\boldsymbol\mu\left(\mathbf{r}\right)\int_\Omega { 
	    \boldsymbol\theta^\top\boldsymbol\lambda\left(\mathbf{r}-\mathbf{r}'\right)
	    \boldsymbol\mu^\top\left(\mathbf{r}'\right) 
	\, \mathrm{d}\mathbf{r}'\mathrm{d}\mathbf{r}} \mathbf{x}_t
	+ \mathbf{w}_t.
\end{equation}
Now we can further simplify the system by defining
\begin{align}
	\label{eq:Lambdatheta}
	 \mathbf{\Lambda}_{\theta} &\triangleq \int_{\Omega}\boldsymbol\mu\left(r\right) \int_\Omega { 
		   \boldsymbol\theta^\top\boldsymbol\lambda\left(r-r'\right)
		    \boldsymbol\mu^\top\left(r\right)\ \mathrm{d}r'\mathrm{d}r},
\end{align}
to give the state equation
\begin{equation}\label{eq:StateEquation}
 \mathbf x_{t+1} =\mathbf A(\boldsymbol \theta) \mathbf x_t+ \mathbf w_t,
\end{equation} 
with the parameterized transition matrix
\begin{equation}\label{eq:A_theta}
 \mathbf A(\boldsymbol \theta)= T_s\varsigma\mathbf{\Lambda}_{x}^{-1}\mathbf{\Lambda}_{\theta}+\xi\mathbf I.
\end{equation} 
The problem of joint state and parameter estimation for the state-space model given by Eq.~\eqref{eq:StateEquation} and Eq.~\eqref{eq:ReducedObservationEquation} can now be formulated as 
\begin{equation}
	{\hat{\mathbf X},\hat{\boldsymbol\theta}}=\arg\max_{\mathbf X,\boldsymbol\theta}p(\mathbf Y;\boldsymbol\theta),
 \end{equation}  
where $\mathbf X$ and $\mathbf Y$ denote the entire sequences of the states and observations respectively, $\boldsymbol\theta$ is the parameter set and $p(\mathbf Y;\boldsymbol\theta)$ denotes the probability density function of the observations parameterized  by $\boldsymbol\theta$, known as the likelihood function. This dual estimation problem cannot be solved separately, as for the state estimation the knowledge of the system's parameters is required and for the parameter estimation the states of the system need to be known. One approach is to iteratively estimate the states and parameters of the system and monitor the convergence of the algorithm. A natural framework to perform this is the well known Expectation Maximization (EM) algorithm \citep{Dempster1977,Shumway2000} to infer both the parameter set and the states from the observations corresponding to the connectivity kernel, the neural field and electrophysiological data in our application. The EM algorithm is applicable  in two main situations:  first is when the observations are incomplete due to a faulty observation process or existing limitations and the second is when the direct maximization of the likelihood function is difficult but can be simplified by assuming the existence of an additional data referred to as hidden data \citep{Bilmes1998}, the weights of the field basis functions in this case. The EM algorithm, when used in the latter context \citep{Dewar2009}, yields the maximum likelihood kernel estimate, i.e.
\begin{equation}
	\boldsymbol\theta_{\text{ML}}=\arg\max_{\boldsymbol\theta}~p(\mathbf Y;\boldsymbol\theta),
 \end{equation}   
and the posterior distribution of the field over time.

@@

For the linear state-space model given by Eqs.~\eqref{eq:StateEquation} and \eqref{eq:ReducedObservationEquation}, the bound defined by Eq.~\eqref{eq:Bishopbound} is a quadratic of the form (see \ref{ap:QDerivation} for derivation) 

@@

where $[\cdot]_{i:} $ denotes the $i^{th}$ row. The matrices $\boldsymbol\Xi_0$ and $\boldsymbol\Xi_1$ in Eqs.~\eqref{eq:epsilon0}-\eqref{eq:Epsilon} are calculated using the RTS smoother outputs: $\hat{\mathbf x}_t$, covariance, $\mathbf P_t^b$, and cross-covariance matrix, $\mathbf M_t$ \citep{Gibsona2005} (see \ref{ap:Xiderivation} for details), 

@@

\boldsymbol\Xi_0&=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t\mathbf{\hat x}_{t+1}^\top\right) \label{eq:Xi0} \\
 \boldsymbol\Xi_1&=\sum_{t=0}^{T-1}\left(\mathbf P_t+\mathbf{\hat x}_t\mathbf{\hat x}_t^\top\right).  \label{eq:Xi1}

@@

\begin{align}\label{eq:Xi0derivation0} 
 \mathbf M_{t+1}&=\mathbf E_{\theta'}\left[(\mathbf x_t-\mathbf{\hat x}_t)(\mathbf x_{t+1}-\mathbf{\hat x}_{t+1})^\top\right] \nonumber \\
&=\mathbf E_{\theta'}\left[\mathbf x_t\mathbf x_{t+1}^\top\right]-\mathbf E_{\theta'}\left[\mathbf x_t\hat{\mathbf x}_{t+1}^\top\right]-\mathbf E_{\theta'}\left[\hat{\mathbf x}_t\mathbf x_{t+1}^\top\right]+\mathbf E_{\theta'}\left[\hat{\mathbf x}_t\hat{\mathbf x}_{t+1}^\top\right]  \nonumber \\
&= \mathbf E_{\theta'}\left[\mathbf x_t\mathbf x_{t+1}^\top\right]-\hat{\mathbf x}_t\hat{\mathbf x}_{t+1}^\top-\hat{\mathbf x}_t\hat{\mathbf x}_{t+1}^\top+\hat{\mathbf x}_t\hat{\mathbf x}_{t+1}^\top\nonumber \\
&= \mathbf E_{\theta'}\left[\mathbf x_t\mathbf x_{t+1}^\top\right]-\hat{\mathbf x}_t\hat{\mathbf x}_{t+1}^\top.
\end{align}
Rearranging   Eq.~\eqref{eq:Xi0derivation0} gives
\begin{equation}\label{eq:Xi0derivation1}
 \mathbf E_{\theta'}\left[\mathbf x_t\mathbf x_{t+1}^\top\right]=\mathbf M_{t+1}+\mathbf {\hat x}_t\mathbf{\hat x}_{t+1}^\top.
\end{equation}
Substituting Eq.~\eqref{eq:Xi0derivation1} into Eq.~\eqref{eq:appXi0}, noting the expectation commutes with the sum yields
\begin{equation}
 \boldsymbol\Xi_0=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t\mathbf{\hat x}_{t+1}^\top\right)
\end{equation}   
An expression for $\boldsymbol\Xi_1$ in terms of the smoothed states can be also found in a similar way by calculating the covariance matrix, $\mathbf P_t$ and following the above steps.
%\begin{equation}\label{eq:Xi1derivation1}
% \mathbf E_{\theta'}\left[\mathbf x_{t}\mathbf x_{t}^\top\right]=\mathbf P_{t}+\mathbf {\hat x}_{t}\mathbf{\hat x}_{t}^\top,
%\end{equation}
%substituting \ref{eq:Xi1derivation1} into \ref{eq:appXi1} we get
%\begin{equation}
 %\boldsymbol\Xi_1=\sum_{t=0}^{T-1}\left(\mathbf P_{t}+\mathbf{\hat x}_{t}\mathbf{\hat x}_{t}^\top\right).
%\end{equation}
\newpage
\bibliographystyle{./elsarticle/model2-names}  
\bibliography{MRAIDE}
\end{document}

>>>>

Neural field models describe the mass action of the central nervous system and are a critical link in our understanding of the biophysics leading to the generation of  electroencephalography (EEG) and magnetoencephalography (MEG) signals. The key components of these models typically describe the macroscopic dynamics of the human brain, but can also be descriptive of finer-scale neurodynamics. Examples are the post-synaptic response kernels (pulse to wave conversion) and the activation function (wave to pulse conversion) \citep{Jirsa1996,Jirsa1997}. The former can be descriptive of inhibitory and excitatory post-synaptic potentials in populations of microscopic single neural models, such as integrate and fire neurons, or an ensemble response to stimulation in macroscopic neural mass models \citep{Stefanescu2008}. The latter can be used to model spiking statistics across spatial scales from single neurons (where it can be considered a cumulative distribution function (CDF) of the probability of firing) as mean firing rates of neural masses. The key factor linking the functions of neural mass models to particular spatial scales is the connectivity, hence the ability to infer a multi-resolution connectivity kernel will enable the formulation of a multi-resolution neural field model.

@@

In this paper, $4^{th}$ order cardinal B-spline scaling and wavelet functions are used, resulting in  $8^{th}$ and $12^{th}$ order B-spline functions when calculating Eq.~\eqref{eq:Lambdax} and Eq.~\eqref{eq:Uij} respectively  whose values at integer points are given in Table~\ref{table:MRA-BsplineatIntegerPoints}.

@@

The simplifications used in formulating the equations enable the development of new methods for extracting information from electrophysiological data. The trade-off between parsimony and complexity is always difficult when considering neurodynamics and this is further complicated when using estimation algorithms that have huge computational demands with such high-dimensional systems.

@@

By assuming a linear activation function of the form
\begin{equation}
	f(v_t(\mathbf{r})) = \varsigma v_t(\mathbf{r}),
\end{equation}
the state transition equation~\eqref{eq:ApproxDiscreteTimeModel4} simplifies to
\begin{equation}\label{eq:ApproxDiscreteTimeModel_Linear}
	\mathbf{x}_{t+1} = 
	\xi \mathbf{x}_t + 
	\varsigma T_s \mathbf{\Lambda}_{x}^{-1} \int_{\Omega}\boldsymbol\mu\left(\mathbf{r}\right)\int_\Omega { 
	    \boldsymbol\theta^\top\boldsymbol\lambda\left(\mathbf{r}-\mathbf{r}'\right)
	    \boldsymbol\mu^\top\left(\mathbf{r}'\right) 
	\, \mathrm{d}\mathbf{r}'\mathrm{d}\mathbf{r}} \mathbf{x}_t
	+ \mathbf{w}_t.
\end{equation}
Now we can further simplify the system by defining
\begin{align}
	\label{eq:Lambdatheta}
	 \mathbf{\Lambda}_{\theta} &\triangleq \int_{\Omega}\boldsymbol\mu\left(r\right) \int_\Omega { 
		   \boldsymbol\theta^\top\boldsymbol\lambda\left(r-r'\right)
		    \boldsymbol\mu^\top\left(r\right)\ \mathrm{d}r'\mathrm{d}r},
\end{align}
to give the state equation
\begin{equation}\label{eq:StateEquation}
 \mathbf x_{t+1} =\mathbf A(\boldsymbol \theta) \mathbf x_t+ \mathbf w_t,
\end{equation} 
with the parameterized transition matrix
\begin{equation}\label{eq:A_theta}
 \mathbf A(\boldsymbol \theta)= T_s\varsigma\mathbf{\Lambda}_{x}^{-1}\mathbf{\Lambda}_{\theta}+\xi\mathbf I.
\end{equation} 
The problem of joint state and parameter estimation for the state-space model given by Eq.~\eqref{eq:ReducedObservationEquation} and Eq.~\eqref{eq:StateEquation}  can now be formulated as 
\begin{equation}
	{\hat{\mathbf X},\hat{\boldsymbol\theta}}=\arg\max_{\mathbf X,\boldsymbol\theta}p(\mathbf Y;\boldsymbol\theta),
 \end{equation}  
where $\mathbf X$ and $\mathbf Y$ denote the entire sequences of the states and observations respectively, $\boldsymbol\theta$ is the parameter set and $p(\mathbf Y;\boldsymbol\theta)$ denotes the probability density function of the observations parameterized  by $\boldsymbol\theta$, known as the likelihood function. This dual estimation problem cannot be solved separately, as for the state estimation the knowledge of the system's parameters is required and for the parameter estimation the states of the system need to be known. One approach is to iteratively estimate the states and parameters of the system and monitor the convergence of the algorithm. A natural framework to perform this is the well known EM algorithm \citep{Dempster1977,Shumway2000} to infer both the parameter set and the states from the observations corresponding to the connectivity kernel, the neural field and electrophysiological data in our application. The EM algorithm is applicable  in two main situations:  first is when the observations are incomplete due to a faulty observation process or existing limitations and the second is when the direct maximization of the likelihood function is difficult but can be simplified by assuming the existence of an additional data referred to as hidden data \citep{Bilmes1998}, the weights of the field basis functions in this case. The EM algorithm, when used in the latter context \citep{Dewar2009}, yields the maximum likelihood kernel estimate, i.e.
\begin{equation}
	\boldsymbol\theta_{\text{ML}}=\arg\max_{\boldsymbol\theta}~p(\mathbf Y;\boldsymbol\theta),
 \end{equation}   
and the posterior distribution of the field over time.

@@

For the linear state-space model given by Eqs.~\eqref{eq:ReducedObservationEquation} and \eqref{eq:StateEquation}, the bound defined by Eq.~\eqref{eq:Bishopbound} is a quadratic of the form (see \ref{ap:QDerivation} for derivation) 

@@

where $[\cdot]_{i:} $ denotes the $i^{th}$ row. The matrices $\boldsymbol\Xi_0$ and $\boldsymbol\Xi_1$ in Eqs.~\eqref{eq:epsilon0}-\eqref{eq:Epsilon} are calculated using the RTS smoother outputs: $\hat{\mathbf x}_t^b$, covariance, $\mathbf P_t^b$, and cross-covariance matrix, $\mathbf M_t$ \citep{Gibsona2005} (see \ref{ap:Xiderivation} for details), 

@@

\boldsymbol\Xi_0&=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t^b\mathbf{\hat x}_{t+1}^{b\top}\right) \label{eq:Xi0} \\
 \boldsymbol\Xi_1&=\sum_{t=0}^{T-1}\left(\mathbf P_t^b+\mathbf{\hat x}_t^b\mathbf{\hat x}_t^{b\top}\right).  \label{eq:Xi1}

@@

\begin{align}\label{eq:Xi0derivation0} 
 \mathbf M_{t+1}&=\mathbf E_{\theta'}\left[(\mathbf x_t-\mathbf{\hat x}_t^b)(\mathbf x_{t+1}-\mathbf{\hat x}_{t+1}^b)^\top\right] \nonumber \\
&=\mathbf E_{\theta'}\left[\mathbf x_t\mathbf x_{t+1}^\top\right]-\mathbf E_{\theta'}\left[\mathbf x_t\hat{\mathbf x}_{t+1}^{b\top}\right]-\mathbf E_{\theta'}\left[\hat{\mathbf x}_t^b\mathbf x_{t+1}^\top\right]+\mathbf E_{\theta'}\left[\hat{\mathbf x}_t^b\hat{\mathbf x}_{t+1}^{b\top}\right]  \nonumber \\
&= \mathbf E_{\theta'}\left[\mathbf x_t\mathbf x_{t+1}^\top\right]-\hat{\mathbf x}_t^b\hat{\mathbf x}_{t+1}^{b\top}-\hat{\mathbf x}_t^b\hat{\mathbf x}_{t+1}^{b\top}+\hat{\mathbf x}_t^b\hat{\mathbf x}_{t+1}^{b\top}\nonumber \\
&= \mathbf E_{\theta'}\left[\mathbf x_t\mathbf x_{t+1}^\top\right]-\hat{\mathbf x}_t^b\hat{\mathbf x}_{t+1}^{b\top}.
\end{align}
Rearranging   Eq.~\eqref{eq:Xi0derivation0} gives
\begin{equation}\label{eq:Xi0derivation1}
 \mathbf E_{\theta'}\left[\mathbf x_t\mathbf x_{t+1}^\top\right]=\mathbf M_{t+1}+\mathbf {\hat x}_t^b\mathbf{\hat x}_{t+1}^{b\top}.
\end{equation}
Substituting Eq.~\eqref{eq:Xi0derivation1} into Eq.~\eqref{eq:appXi0}, noting the expectation commutes with the sum yields
\begin{equation}
 \boldsymbol\Xi_0=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t^b\mathbf{\hat x}_{t+1}^{b\top}\right)
\end{equation}   
An expression for $\boldsymbol\Xi_1$ in terms of the smoothed states can be also found in a similar way by calculating the covariance matrix, $\mathbf P_t^b$ and following the above steps.
%\begin{equation}\label{eq:Xi1derivation1}
% \mathbf E_{\theta'}\left[\mathbf x_{t}\mathbf x_{t}^\top\right]=\mathbf P_{t}+\mathbf {\hat x}_{t}\mathbf{\hat x}_{t}^\top,
%\end{equation}
%substituting \ref{eq:Xi1derivation1} into \ref{eq:appXi1} we get
%\begin{equation}
 %\boldsymbol\Xi_1=\sum_{t=0}^{T-1}\left(\mathbf P_{t}+\mathbf{\hat x}_{t}\mathbf{\hat x}_{t}^\top\right).
%\end{equation}
\newpage
\bibliographystyle{./elsarticle/model2-names}  
\bibliography{MRAIDE}
\end{document}
